---
title: "Get Started"
output: html_document
---

The assignPOP package is built and used in R, so be sure to have R (and RStudio) pre-installed in your computer.

Download and install R from CRAN [https://cran.r-project.org/](https://cran.r-project.org/) 

Download and install RStudio from [https://www.rstudio.com/](https://www.rstudio.com/)

***
### Install assignPOP to your R environment

Once you have R/RStudio installed in your computer, you can enter the following command in R console to install the package from CRAN.

```{r, eval=FALSE}
install.packages("assignPOP")
```

Or, install from Github site (which requires the package `devtools` pre-installed).

```{r, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("alexkychen/assignPOP")

```

*Note*: When you install the package from Github, you may need to install additional packages before the assignPOP can be successfully installed. Follow the hints that R provided and then re-run `install_github("alexkychen/assignPOP")`.


***
### Import assignPOP to R

```{r, eval=FALSE}
library(assignPOP)

```

***

**The assignPOP package includes the following functions.**

`read.genpop()` Read genetic data in GENEPOP format

`reduce.allele()` Reduce low variance loci

`compile.data()` Concatenate genetic and non-genetic data

`assign.MC()` Perform population assignment using Monte-Carlo cross-validation

`assign.kfold()` Perform population assignment using K-fold cross-validation

`assign.X()` Assign unknown individuals to source populations

`accuracy.MC()` Calculate assignment accuracy from Monte-Carlo cross-validation results

`accuracy.kfold()` Calculate assignment accuracy from K-fold cross-validation results

`accuracy.plot()` Make assignment accuracy box plots

`membership.plot()` Make membership probability stacked bar plots

`check.loci()` Identify informative loci used in resampling cross-validation

**Enter `?` and function name in R console (e.g.,`?read.genpop`) to learn more about the functions and what arguments are used in the function.**

