---
title: "Get Started"
output: html_document
---

If you are an R first time user, please download and install R (and RStudio) in your computer.

Download and install R from CRAN [https://cran.r-project.org/](https://cran.r-project.org/) 

Download and install RStudio from [https://www.rstudio.com/](https://www.rstudio.com/)

***
### Install assignPOP to your R environment

Once you have R/RStudio installed, you can enter the following command in R console to install the package from CRAN.

```{r, eval=FALSE}
install.packages("assignPOP")
```

Or, install via Github site (which requires the package `devtools` pre-installed).

```{r, eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("alexkychen/assignPOP")

```

*Note*: When you install the package via Github, you may need to install additional packages before the assignPOP can be successfully installed. Follow the hints that R provides and then re-run `install_github("alexkychen/assignPOP")`.


***
### Import assignPOP to R

```{r, eval=FALSE}
library(assignPOP)

```

***

####**The assignPOP package includes the following functions.**####

| Function | Return | Description |
|:---------|:-------|:------------|
| `read.Genepop()` | A list object including data matrix | Read genetic data in GENEPOP format |
| `read.Structure()` | A list object including data matrix | Read genetic data in STRUCTURE format |
| `reduce.allele()` | A list object including data matrix | Reduce low variance loci |
| `compile.data()` | A list object including data matrix | Concatenate genetic and non-genetic data |
| `assign.MC()` | Output results in text files | Perform population assignment using Monte-Carlo cross-validation |
| `assign.kfold()` | Output results in text files | Perform population assignment using K-fold cross-validation |
| `assign.X()` | Output results in text files | Assign unknown individuals to source populations |
| `accuracy.MC()` | A data frame and text file | Calculate assignment accuracy from Monte-Carlo cross-validation results |
| `accuracy.kfold()` | A data frame and text file | Calculate assignment accuracy from K-fold cross-validation results |
| `accuracy.plot()` | ggplot2 graph | Make assignment accuracy box plots |
| `membership.plot()` | ggplot2 graph | Make membership probability stacked bar plots |
| `assign.matrix()` | Contingency tables | Print mean and standard deviation across assignment tests |
| `check.loci()` | Output results in a text file | Identify informative loci used in resampling cross-validation |

**Enter `?` followed by a function name in R console (e.g.,`?read.Genepop`) to see the help page.**

